---
title: "Challenge Exercise"
---

This exercise uses models to describe where we have **no data at all**.

* data: male and female incomes at age 30--50 in 2010--2019
* task: what was the male and female geometric mean pay at age 30--50 in **2022**?

Caution is always needed when carrying out model-based extrapolation. But when forecasting, a model is also often the only way to make progress.

You can generate data as on the [Simulate Data](simulate_a_sample) page. You can set the sample size however large you want. You can use any model you want.

You might discuss methodological choices:

* how would you generate an evaluation set for this problem?
* what models do you think would work well?

You might also discuss conceptual issues:

* why might we be hesitant to carry out this extrapolation?

## An example to get you started

As a simple example, I might simulate a sample of size 100,

```{r, echo = F, results = "hide", message = F, warning = F}
library(tidyverse)
simulate <- function(n = 100) {
  read_csv("assets/truth.csv") |>
    slice_sample(n = n, weight_by = weight, replace = T) |>
    mutate(income = exp(rnorm(n(), meanlog, sdlog))) |>
    select(year, age, sex, income)
}
```

```{r, results = "hide"}
simulated <- simulate(n = 100)
```

estimate a linear model on those data,

```{r}
fit <- lm(log(income) ~ sex * year, data = simulated)
```

and report predictions in 2022.
```{r}
to_predict <- tibble(
  sex = c("female","male"),
  year = c(2022,2022)
)
to_predict |>
  mutate(
    # Make prediction
    estimate = predict(fit, newdata = to_predict),
    # Exponentiate to dollars
    estimate = exp(estimate)
  )
```

## What to try next

You might consider different functional forms, the overall mean, or machine learning estimators.
