---
title: "Why model?"
---

We model because

* there is not enough data in our target subgroup
* we have a good idea to share information with other cases

To run the code on this page, you will need the `tidyverse` package.

```{r, message = F, warning = F, results = "hide"}
library(tidyverse)
```

We will also set the seed so that it is possible to exactly reproduce these results.

```{r}
set.seed(90095)
```

## Generate an illustration sample

The code below will generate a sample of 100 respondents simulated to correspond to the target population ages 30--50 in 2010--2019.

```{r, comment = F, message = F}
simulate <- function(n = 100) {
  read_csv("assets/truth.csv") |>
    slice_sample(n = n, weight_by = weight, replace = T) |>
    mutate(income = exp(rnorm(n(), meanlog, sdlog))) |>
    select(year, age, sex, income)
}
```

```{r, message = F}
simulated <- simulate(n = 100)
```
```{r, echo = F}
num_female <- simulated |>
  filter(age == 30 & sex == "female") |>
  nrow()
num_male <- simulated |>
  filter(age == 30 & sex == "male") |>
  nrow()
while (num_female == 0 | num_male == 0) {
  simulated <- simulate(n = 100)
  num_female <- simulated |>
    filter(age == 30 & sex == "female") |>
    nrow()
  num_male <- simulated |>
    filter(age == 30 & sex == "male") |>
    nrow()
}
simulated |> print(n = 3)
```

## Target estimand

We will estimate three target quantities:

* (1) geometric mean pay among female respondents age 30
* (2) geometric mean pay among male respondents age 30
* (3) ratio (1) / (2)

In our sample, there are only `r num_female` female and `r num_male` male 30-year-olds! We will need a model.

```{r, eval = F, echo = F}
simulated |>
  filter(age == 30 & sex == "female")
```

## Models for illustration

We consider a series of three models.

* Flat model: Geometric mean among everyone
* Linear model: Prediction from linear fit on age $\times$ sex
* Quadratic model: Prediction from quadratic fit on age $\times$ sex

Recall that each model is a tool to share information from other cases in order to predict an estimate for the target subgroups: 30-year-old male and female respondents.

```{r, echo = F}
truth <- read_csv("assets/truth.csv") |>
  group_by(sex, age) |>
  summarize(meanlog = weighted.mean(meanlog, w = weight))

estimators <- function(
    data,
    to_predict = tibble(
      sex = rep(c("female","male"),2),
      age = rep(c(30,45),each = 2)
    )
) {
  fit_flat <- lm(log(income) ~ sex, data = data)
  fit_linear <- lm(log(income) ~ sex*age, data = data)
  fit_quadratic <- lm(log(income) ~ sex*poly(age,2), data = data)
  
  to_predict |>
    left_join(
      data |> group_by(age, sex) |> summarize(point = exp(mean(log(income))), .groups = "drop"),
      by = join_by(age, sex)
    ) |>
    mutate(
      flat = exp(predict(fit_flat, newdata = to_predict)),
      linear = exp(predict(fit_linear, newdata = to_predict)),
      quadratic = exp(predict(fit_quadratic, newdata = to_predict))
    ) |>
    pivot_longer(cols = c("point","flat","linear","quadratic"), names_to = "estimator", values_to = "estimate") |>
    # Calculate the female / male ratio of estimates
    pivot_wider(names_from = "sex", values_from = "estimate") |>
    mutate(ratio = female / male) |>
    pivot_longer(
      cols = c("female","male","ratio"), 
      names_to = "estimand", 
      values_to = "estimate"
    ) |>
    mutate(sample_size = nrow(data))
}

estimators(
  data = truth |> mutate(income = exp(meanlog)),
  to_predict = tibble(
      sex = rep(c("female","male"),16),
      age = rep(c(30:45),each = 2)
    )
) |>
  left_join(truth |> mutate(truth = exp(meanlog)) |> rename(estimand = sex), by = c("estimand","age")) |>
  filter(estimand != "ratio") |>
  filter(estimator != "point") |>
  ggplot(aes(x = age, color = estimand)) +
  geom_line(aes(y = estimate)) +
  geom_point(aes(y = truth, shape = age == 30)) +
  facet_wrap(
    ~estimator,
    labeller = as_labeller(function(x) {
      case_when(
        x == "flat" ~ "Flat Model",
        x == "linear" ~ "Linear Model",
        x == "quadratic" ~ "Quadratic Model"
      )
    })
  ) + 
  ggtitle("Illustrated on full data: 5 million cases") +
  scale_color_discrete(
    name = "Sex Subgroup",
    labels = c("Female","Male")
  ) +
  scale_x_continuous(name = "Age Subgroup") +
  scale_y_continuous(name = "Geometric Mean Income\n(Log Scale)", labels = scales::label_dollar(), trans = "log") +
  scale_shape_manual(
    name = element_blank(),
    labels = function(x) ifelse(x,"Target Subgroup","Other Cases"),
    values = c(1,16)
  )
ggsave("slides/define_models.pdf", height = 3, width = 7)
```

```{r, echo = F, eval = F}
# This is just a figure for slides
truth |> 
  mutate(truth = exp(meanlog)) |> 
  rename(estimand = sex) |>
  ggplot(aes(x = age, color = estimand)) +
  geom_point(aes(y = truth, shape = age == 30)) +
  ggtitle("Illustrated on full data: 5 million cases") +
  scale_color_discrete(
    name = "Sex Subgroup",
    labels = c("Female","Male")
  ) +
  scale_x_continuous(name = "Age Subgroup") +
  scale_y_continuous(name = "Geometric Mean Income\n(Log Scale)", labels = scales::label_dollar(), trans = "log") +
  scale_shape_manual(
    name = element_blank(),
    labels = function(x) ifelse(x,"Target Subgroup","Other Cases"),
    values = c(1,16)
  )
ggsave("slides/true_age_pattern.pdf", height = 3, width = 5)
```

We will use these models for a task: estimate the geometric mean pay among 30-year-old male and female respondents. We will carry out the task in a simulated sample of 100 respondents.

## Which model would you choose?

Think about an answer before going on to the next page.

