---
title: "Simulate Data"
---

This exercise works with simulated samples. Taking the [nonparametric estimates](assets/truth.csv) from 5 million cases as the truth, you will generate a simulated sample of a much smaller size using the code below.

If you are a Stata user, see the bottom of this page for code. The page mainly supports coding in R.

Prepare the environment by loading the `tidyverse` package.

```{r, message = F, warning = F}
library(tidyverse)
```

The function below simulates a sample of 100 cases.

```{r, echo = F}
simulate <- function(n = 100) {
  read_csv("assets/truth.csv") |>
    slice_sample(n = n, weight_by = weight, replace = T) |>
    mutate(income = exp(rnorm(n(), meanlog, sdlog))) |>
    select(year, age, sex, income)
}
```

```{r, eval = F}
simulate <- function(n = 100) {
  read_csv("https://ilundberg.github.io/description/assets/truth.csv") |>
    slice_sample(n = n, weight_by = weight, replace = T) |>
    mutate(income = exp(rnorm(n(), meanlog, sdlog))) |>
    select(year, age, sex, income)
}
```

We can see how it works below,

```{r, results = 'hide'}
simulated <- simulate(n = 100)
```

and can print a bit of the output.

```{r}
simulated |> print(n = 3)
```

## Code for Stata users

I am mostly not a Stata user, and this is provided for secondary pedadogical purposes in case some people do not use R. If you are a Stata user, feel free to let me know how to improve this code.

```{r, eval = F}
set seed 90095

* Load true population data

import delimited https://ilundberg.github.io/description/assets/truth.csv

* Draw a sample of 100 X-values
* Need two supporting packages
*ssc install moremata
*ssc install gsample

* Draw the sample
gsample 100 [w = weight]

* Simulate individual income data
gen log_income = meanlog + sdlog * rnormal()
gen income = exp(log_income)

* Keep variables to work with

encode sex, gen(factorsex)
keep year age factorsex log_income income
rename factorsex sex
```